Type Evolution in Interactive Programming Systems

### 2024-05-31 paper.tex
Spellecheck

% Approach: What was done that unveiled new knowledge?
In this paper, we develop a unfied conceptual framework for analysing the problem of type evolution.
We use the framework to study type evolution in five diverse case studies.
%
% Knowledge: What new facts were uncovered? What new capabilities are enabled by the work?
By looking at the case studies through a unified perspective, we hope to inspire new thinking
about the problem of type evolution and also encourage knowledge transfer across sub-fields.

@@

Looking at type evolution from the perspective of programming systems can offer a new perspective that
unfies the many different aspects of the problem partially explored in other sub-fields. A programming
system may have permanent data (as in a database) whose structure can change, requiring
corresponding change to data and code. A programming system may have a transient state conforming
to a type structure that may change during live programming. A programming system may also need
to allow programmers to independently work on different versions of a program.

@@

More generally, there is a number of edits to the types that the programmer may want to do
without restarting the application. The edits and the desired handling in the programming
system are summarized in Table~\ref{tbl:elmchanges}. Modifying the \texttt{Event} type does not
require data migration (assuming that the change does not happen when there are unprocessed
events). It may result in unused code or missing cases in \texttt{update} that need to be addressed.
Modifying \texttt{State} ranges from relatively simple problems (e.g., adding a new field with a
default value) to challenging case. In particular, if the programer refactors \texttt{State} to a
semantically equivalent type, the programming system should, in principle, be able to automatically
migrate both the original value and implementation of both functions to match the new type.

@@

\challenge{Extract Entity}
Acme needs to evolve the above schema and data into two tables (Fig.~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
one with orders that links to one with customers.  We refer to this as the \emph{Extract Entity}
operation, because it introduces a new \emph{entity}, a referenceable holder of mutable attributes.
The operation is needed when we realize that some attributes of one type of entity should belong to
a  distinct type of entity that will be associated with the first one.

@@

Entity evolution can happen through operations such as those listed in
Table~\ref{tbl:dbchanges}. Every schema operation should have an inverse as the
change in requirements that triggered an evolution might change back again.
Invertibility features in other work on schema evolution and \emph{bidirectional transformations}~\cite{Foster2007, herrmann17, chillon22}.
The inverse of \emph{Extract Entity} is \emph{Absorb Entity}. It raises the question of how
to handle \emph{one-to-many} relationships. For example should absorbing orders into customers
do a relational join, returning us back to the starting point, or should it nest orders within
customers, as a NoSQL database might do?

@@

The challenge focuses on collaborative conference planning, illustred in Figure~\ref{fig:conf}
and inspired by a typical use of Notion~\cite{notion}. A number of co-organizers are planning
a conference that will feature talks by several invited speakers. They need to agree on the
speaker list, contact speakers and calculate the conference budget.
%
The ideas in this section also apply to work on computational scientific notebooks
that combine code and data such as Jupyter \cite{Kluyver2016}. We hope to contribute to ongoinig
work on advancing the format as done by Nextjournal \cite{Nextjournal21}, Nota~\cite{Crichton2021}
and Living Papers~\cite{Heer2023}.

@@

\subsection*{Remarks: Liveness and Divergence}
The above challenge focuses on the local-first software scenario, but it can be extended to also
apply to live programming. If formulas in the document are evaluated on-the-fly, the system
needs to identify edits that invalide some of the values (by changing the data that a formula
depends on, or by changing the equation). It should then automatically re-evaluate or invalidate
the computed results (requiring explicit re-evaluation). In this scenario, we distinguish between
\emph{permanent data} (information in the document) and transient \emph{data} (computed by code).
Whereas permanent data should be synchronized and co-evolve with the document structure (type),
transient data can be discarded.

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-1.5em}
\includegraphics[width=14em]{figures/arr-backport.png}
\vspace{-2.5em}
\end{wrapfigure}
\frameworkboxtitle{Bidirectional Data Migration}
The scenario adds the need for transferring data changes from a new type to an older type.
In the diagram on the right, we only show data and type, ignoring the code added by the
accounting department. We start with data $D_1$ of type~$T_1$. The accounting deparment
evolves the type (via Extract Entity), obtaining data $D_2$ of type $T_2$ using an automatically
generated forward transformation. It further modifies the data, obtiaing $D_3$. When the
accounting department receives data $D_3$, it needs to migrate it back from type $T_2$ to
$T_1$ using an automatically generated backward transformation. That is,
type evolution needs to be accompanied with both forward and backward projection for data
(and code) migration.}

>>>>

% Approach: What was done that unveiled new knowledge?
In this paper, we develop a unified conceptual framework for analysing the problem of type evolution.
We use the framework to study type evolution in five diverse case studies.
%
% Knowledge: What new facts were uncovered? What new capabilities are enabled by the work?
By looking at the case studies through a unified perspective, we hope to inspire new thinking
about the problem of type evolution and also encourage knowledge transfer across sub-fields.

@@

Looking at type evolution from the perspective of programming systems can offer a new perspective that
unifies the many different aspects of the problem partially explored in other sub-fields. A programming
system may have permanent data (as in a database) whose structure can change, requiring
corresponding change to data and code. A programming system may have a transient state conforming
to a type structure that may change during live programming. A programming system may also need
to allow programmers to independently work on different versions of a program.

@@

More generally, there is a number of edits to the types that the programmer may want to do
without restarting the application. The edits and the desired handling in the programming
system are summarized in Table~\ref{tbl:elmchanges}. Modifying the \texttt{Event} type does not
require data migration (assuming that the change does not happen when there are unprocessed
events). It may result in unused code or missing cases in \texttt{update} that need to be addressed.
Modifying \texttt{State} ranges from relatively simple problems (e.g., adding a new field with a
default value) to challenging case. In particular, if the programmer refactors \texttt{State} to a
semantically equivalent type, the programming system should, in principle, be able to automatically
migrate both the original value and implementation of both functions to match the new type.

@@

\challenge{Extract Entity}
Acme needs to evolve the above schema and data into two tables (Fig.~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
one with orders that links to one with customers.  We refer to this as the \emph{Extract Entity}
operation, because it introduces a new \emph{entity}, a referencable holder of mutable attributes.
The operation is needed when we realize that some attributes of one type of entity should belong to
a  distinct type of entity that will be associated with the first one.

@@

Entity evolution can happen through operations such as those listed in
Table~\ref{tbl:dbchanges}. Every schema operation should have an inverse as the
change in requirements that triggered an evolution might change back again.
Invertability features in other work on schema evolution and \emph{bidirectional transformations}~\cite{Foster2007, herrmann17, chillon22}.
The inverse of \emph{Extract Entity} is \emph{Absorb Entity}. It raises the question of how
to handle \emph{one-to-many} relationships. For example should absorbing orders into customers
do a relational join, returning us back to the starting point, or should it nest orders within
customers, as a NoSQL database might do?

@@

The challenge focuses on collaborative conference planning, illustrated in Figure~\ref{fig:conf}
and inspired by a typical use of Notion~\cite{notion}. A number of co-organizers are planning
a conference that will feature talks by several invited speakers. They need to agree on the
speaker list, contact speakers and calculate the conference budget.
%
The ideas in this section also apply to work on computational scientific notebooks
that combine code and data such as Jupyter \cite{Kluyver2016}. We hope to contribute to ongoing
work on advancing the format as done by Nextjournal \cite{Nextjournal21}, Nota~\cite{Crichton2021}
and Living Papers~\cite{Heer2023}.

@@

\subsection*{Remarks: Liveness and Divergence}
The above challenge focuses on the local-first software scenario, but it can be extended to also
apply to live programming. If formulas in the document are evaluated on-the-fly, the system
needs to identify edits that invalidate some of the values (by changing the data that a formula
depends on, or by changing the equation). It should then automatically re-evaluate or invalidate
the computed results (requiring explicit re-evaluation). In this scenario, we distinguish between
\emph{permanent data} (information in the document) and transient \emph{data} (computed by code).
Whereas permanent data should be synchronized and co-evolve with the document structure (type),
transient data can be discarded.

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{-1.5em}
\includegraphics[width=14em]{figures/arr-backport.png}
\vspace{-2.5em}
\end{wrapfigure}
\frameworkboxtitle{Bidirectional Data Migration}
The scenario adds the need for transferring data changes from a new type to an older type.
In the diagram on the right, we only show data and type, ignoring the code added by the
accounting department. We start with data $D_1$ of type~$T_1$. The accounting department
evolves the type (via Extract Entity), obtaining data $D_2$ of type $T_2$ using an automatically
generated forward transformation. It further modifies the data, obtaining $D_3$. When the
accounting department receives data $D_3$, it needs to migrate it back from type $T_2$ to
$T_1$ using an automatically generated backward transformation. That is,
type evolution needs to be accompanied with both forward and backward projection for data
(and code) migration.}
